Forecasting the old-age dependency ratio to determine the best pension age

### 2019-01-06 PensionAge.tex
PensionAge.pdf

---
title: "Forecasting the old-age dependency ratio to determine the best pension age"
author:
- familyname: Hyndman
  othernames: Rob J
  address: Department of Econometrics & Business Statistics\newline Monash University, Clayton VIC 3800, Australia
  email: Rob.Hyndman@monash.edu
  correspondingauthor: true
- familyname: Zeng
  othernames: Yijun
  address: Monash University
- familyname: Shang
  othernames: Han Lin
  address: Australian National University
abstract: "We forecast the old-age dependency ratio for Australia under various pension age proposals, and estimate a pension age scheme that will provide a stable old-age dependency ratio at a specified level. Our approach involves a stochastic population forecasting method based on coherent functional data models for mortality, fertility and net migration, which we use to simulate future age-structures of the population. We apply these ideas to Australian historical demographic data and the results suggest that the pension age should be increased to 67 by 2021, 70 by 2032 and 75 by 2058 in order to maintain the old-age dependency ratio at the current level, which is 21%. This approach can easily be extended to other target levels of the old-aged dependency ratio and to other countries."
keywords: "functional time series model, coherent forecasts, demographic components, pension and insurance"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: pensionage.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages = FALSE, warning = FALSE, cache = TRUE)
# Make sure you have the latest version of rmarkdown and bookdown
# devtools::install_github("rstudio/rmarkdown")
# devtools::install_github("rstudio/bookdown")
# And the latest versions of Our packages
# devtools::install_github("robjhyndman/forecast")
# devtools::install_github("robjhyndman/demography")
# devtools::install_github("robjhyndman/addb")
library(demography)
library(addb)
library(ggplot2)
source("functions-v4.R")
```

# Introduction

Australia's ageing population is a result of three factors: (1) significant technological advances in medical care leading to longer life spans; (2) high fertility rates after World War II resulting in the "baby-boomer" generation; and (3) a large decline in fertility rates over the past twenty or so years [@Fehr2008].  Consequently, there are many older people leaving the workforce, and fewer workers replacing them. Similar demographic challenges face many other countries. One way of measuring this problem is via the old-age dependency ratio (OADR): the number of people post-retirement to the number of people of working age [@Walker1982]. The OADR has been increasing for many decades, but recently the rate of increase has jumped markedly, and we show that it will continue to increase at a rapid rate if there are no age-related policy changes.

This is consistent with the Australian Intergenerational Reports [@IGR02; @IGR07; @Swan2010] that explored the extent to which existing age-related policies would affect the burden of future working generations. To date, the Australian government has addressed the problem through a variety of approaches including modifications to compulsory superannuation policies, instigating work incentives for older people, and increasing the pension age (defined as the age at which eligible residents can access the age pension for the first time). In 2009, the federal government announced an increase in pension age from 65 to 67 by 2023 [@Nielson2010]. In May 2014, the Federal Treasurer proposed a further increase in the pension age to 70 by 2035. This new proposal is in line with the recommendations from the @PC13 and the @NCOA2014. In this paper, we investigate the effect of these changes in the pension age on the forecast OADR.

@@

In this paper, we use the OADR as a measure of the financial burden on the Australian workforce because a rise in the OADR will reduce the ability of working age population to finance pensions and health costs [@ARW07; @IGR02; @IGR07].  The OADR is defined as the ratio of the number of people aged over the pension age to the number of people of working age [@ARW07]. This can be expressed using the formula below:
\begin{equation}
  \text{OADR} = \frac{\text{number of people aged over pension age}}{\text{number of people aged 15 to pension age}}*100\%
\end{equation}
We provide forecasts of future population age structures over the next 50 years. We then use those forecasts to compute the OADR associated with pension age both before and after the adjustments. One contribution of the current study is the combined use of the Hyndman-Ullah forecasting method [@HU07] and the product-ratio method [@HBY13] to forecast a population age structure.  Forecasts of population structure have been built on forecasts of age- and gender-specific mortality rates, fertility rates and net migration using this method. These forecasts presuppose that the historical trends of past years will continue. Historical data of those rates have thus been used as input to simulate future paths of age- and gender-specific population structure.

Using these predicted population paths, we can calculate the average OADR and predict the levels of old-age dependency that correspond to different pension age schemes. By testing different pension age schemes, we aim to find the pension age with the desired OADR outcome. Therefore, other contributions of this research are (1) to examine the effect of adjusting pension ages by comparing OADR before and after adjustment; and (2) to suggest a suitable pension age scheme to ensure that the economic burden will remain relatively constant at a desired level in the future. It should be noted that the desired level has been set at 21%, the estimated level of OADR in 2014. Whether this OADR level is the most appropriate level for Australia remains to be examined and is not the focus of this paper. This paper concentrates on providing a framework for policy makers to find a target pension age for a given desired level of OADR.

Some may argue that using OADR as a measure of financial burden for the next generation could be biased, with a large number of retirees able to rely on self-financed retirement income, such as superannuation. However, 80% of Australians over pension age rely fully or partly on the age pension [@Power14]. Therefore, the proportion of people over working age is still a good representation of financial burden on future working generations.  Nevertheless, using OADR as a single measurement of financial burden may overlook some important economic factors, and @HY12 have suggested the use of an adjusted OADR that takes account of some of these factors in analyzing pension finance. We do not consider these here. Our approach is intended to illustrate how population forecasting can be used to explore the pension age issue in a simple context, and it could be extended to take into account other variables.

The remainder of this paper proceeds as follows. Section \@ref(sec:litreview) reviews key literature on population forecasting and the pension age in Australia. Section \@ref(sec:data) describes the data and modelling framework we use in this study.  We show how our framework can be used to estimate a pension age scheme that leads to a target OADR in Section \@ref(sec:target). Finally, Section \@ref(sec:conclusions) gives concluding remarks as well as identifies the scope for future research.

# Background and Literature Review {#sec:litreview}

## The age pension in Australia

@@

Currently, the age pension is the primary source of income for 70% -80% of Australian residents over 64 years [@ABS06]. Even though the government has promoted self-financed retirement by introducing the Superannuation Guarantee in 1992, the Superannuation Guarantee accumulations are still modest, with the current average less than \$100,000 at retirement [@ASFA07].  It has been estimated by @PT08 that a retiree will need to have \$450,000 at 65 in order to support a pension-equivalent income and this is more than eight times of the average retirement accumulation of Australians.  The required wealth to comply with personal-funded retirement savings requirements is too high for most Australians, suggesting that retirees will continue to rely on the age pension in the coming decades.

The ageing trend in Australia, along with retirees' dependence on the age pension, has dramatically increased the fiscal cost of the age pension. For instance, the fiscal cost of the age pension is currently around at 2.5% of Gross Domestic Profit (GDP), but is expected to increase to 4.4% by the 2050s (Australian Government, 2007). The @PC13 also indicates the government will need to raise taxes by 21% to pay for extra aged care and health costs, which is caused by longer life expectancies in newer generations.

The issue is not only a problem for the government, but also for future retirees. Given children born today are expected to live up to their 90s on average, the @PC13 suggests that Australians won't be able to afford to spend 35 years in retirement (i.e., retire around 65). They have predicted that a third of baby boomers are expected to spend all their money and assets before they die\footnote{Baby Boomers are people born during the Post-World War II baby boom between the years 1946 and 1964}. This problem might eventually deteriorate the sustainability of the age pension, although the welfare system works well at the moment, because demographic changes often have delayed effects [@IGR02].

```{r changespension, echo=FALSE, fig.cap="Government's adjustments on pension age scheme.", fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
year <- 2065 - 2007
proposed <- ts(pmin(
  70,
  c(rep(65, 2017 - 2008), 65 + cumsum(rep(c(0.5, 0), 55)))
)[1:50],
start = 2008
)
approved <- pmin(proposed, 67)
current <- pmin(proposed, 65)
cols <- c(current = "#1b9e77", approved = "#d95f02", proposed = "#7570b3")
ggplot() +
  forecast::autolayer(proposed, series = "proposed") +
  forecast::autolayer(approved, series = "approved") +
  forecast::autolayer(current, series = "current") +
  xlab("Year") + ylab("Age of retirement") +
  scale_color_manual(
    values = cols, name = " ",
    breaks = c("current", "approved", "proposed"),
    labels = c("Pension age at 65", "Approved change", "Proposed further change")
  )
```

The Australian government is trying to take early intervention in relation to the pension age to balance between providing social support and fiscal cost. The pension age for males has been set at 65 for over a century. While the pension age for females was 60 years until 1994, it has steadily increased to 65 years, in line with that of males. The previous Labor government announced in 2009 that the pension age will increase six months every two years until 2017, capping at 67 in July 2023. This change is expected to have a significant impact on labor force participation and to decrease the retirement duration of residents because pension age is regarded as a normal age for retirement in Australia. However, the @PC13 proposed further lifting of the pension age to 70 on the grounds that increasing the pension age to 70 would save taxpayers \$150 billion in welfare costs and health spending, and elderly people are capable of continuing work. The National Commission of Audit (NCOA) (2014) holds the same opinion, and recommended lifting the pension age to 70 from 2053. The current Treasurer Joe Hockey announced that the government would increase the pension age to 70 from 2035, which is proposed in the Federal Budget Statement for 2014-2015 to be confirmed by Australian Federal Government. Figure \ref{fig:changespension} illustrates the process of the adjustments on pension age scheme. This announcement received overwhelmingly negative responses in the media. @Guest14 argues from an economic prospective that Australia currently has much better circumstances than other OECD countries in terms of age pension pressure, because spending on the age pension in Australia is  less than half of the OECD average. He also claims that Australia has a higher consumption of all other goods and services, and so why are they not allowed to have more leisure time. Further, @Power14 contends that many Australians are not physically able to work full-time to 70, so that recommendations made by the PC and NCOA are simplistic solutions to a more a complex social problem.

In fact, all supportive arguments for increasing the pension age in Australia have only justified the need to increase the pension age. No justification on the pace or degree of the change has been provided. Also, the government has not disclosed the method it used to determine the new pension age. It is therefore not surprising that the adjustment of the pension age has been questioned by the public. A consistent method to determine the appropriate pension age would be more persuasive in general. For example, it is preferred by OECD countries that pension age should be linked to life expectancy [@OECD12]. There is a practical issue as to how to link life expectancy with pension ages given that there is no information on future mortality rates [@PC13]. @CEDA07 [p7] proposed directly linking pension age with life expectancy in the way that "pension age should increase by approximately 50% of any increase in life expectancy". However, using a "rule of thumb" approach might not be ideal, as 50% of the change in life expectancy may not be a good adjustment to pension age in the long term. Our study suggests an approach, linking the pension age with the the aged dependency ratio, based on forecasts of future mortality and fertility rates. Our approach has the advantage that it can be used to justify the change in pension age and directly links the pension age with our ultimate goal of a stable financial burden for the public.

## Population forecasting

In population forecasting, two very different approaches: a deterministic approach (sometimes called a quasi-statistical approach) and a stochastic approach (a purely statistical approach) are frequently used. For instance, the Australian Bureau of Statistics (ABS) is using a deterministic approach for its demographic projection, while some other countries or institutes forecast demographic trends based on purely statistical approaches such as Statistics New Zealand and the United States Census Bureau. In the deterministic approach, experts project future demographic changes using judgement-based scenarios. It ignores the non-demographic factors such as major pandemics or wars, although historically they are important factors that affect demographic behaviour [@PC13]. In stochastic approaches, the Lee-Carter (LC) method [@LC92] is the most prominent demographic forecasting method, which now has several variants and extensions that relax some of the assumptions of the basic Lee-Carter method [see, e.g., @SBH11]. Although statistical forecasting models are rigorous, they still leave a basis for judgment-based adaptation, as they sometimes produce unrealistic projections of demographic factors such as long-run projections of mortality rates [@PC13].

Therefore, some official agencies decide to mix the two approaches in their forecasts in order to provide a supplement to both approaches. For instance, the most recent forecast of population in Australia by the @PC13 was based on basic Lee-Carter methods plus some judgement-based adaptations to address the characteristics of demographic trends that the Lee-Carter method failed to capture. The approach of forecasting used in this study is very similar to the @PC13. The Hyndman-Ullah method (2007) is used as a basic forecasting approach, which is a comprehensive extension of Lee-Carter method. In addition, an adapted product-ratio method is used to eliminate the potential long-term divergence between female and male mortality rates. A more detailed review of forecasting methods is given below.

### Deterministic Methods Used by the ABS

>>>>

\documentclass[11pt,a4paper,]{article}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Forecasting the old-age dependency ratio to determine the best pension age},
            pdfkeywords={functional time series model, coherent forecasts, demographic
components, pension and insurance},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{pensionage.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Forecasting the old-age dependency ratio to determine the best pension
age}

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\RequirePackage{bera}
\RequirePackage{mathpazo}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*5}{*1}
\titlespacing{\subsection}{0pt}{*2}{*0.2}

%% TITLE PAGE
\def\Date{\number\day}
\def\Month{\ifcase\month\or
 January\or February\or March\or April\or May\or June\or
 July\or August\or September\or October\or November\or December\fi}
\def\Year{\number\year}

\makeatletter
\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
\def\nojel{\def\showjel{}}
\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
\def\cover{{\sffamily\setcounter{page}{0}
        \thispagestyle{empty}%
        \vspace*{-2cm}
        \centerline{\raisebox{-1.8cm}{\includegraphics[width=5cm]{MBSportrait}}\hspace*{9cm} ISSN 1440-771X}\vspace{0.99cm}
        \begin{center}\Large
        Department of Econometrics and Business Statistics\\[.5cm]
        \scriptsize http://business.monash.edu/econometrics-and-business-statistics/research/publications
        \end{center}\vspace{2cm}
        \begin{center}
        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
                \LARGE\@author\end{onehalfspace}
        }}
        \end{center}
        \vfill
                \begin{center}\Large
                \Month~\Year\\[1cm]
                Working Paper \@wp
        \end{center}}}
\def\pageone{{\sffamily\setstretch{1}%
        \thispagestyle{empty}%
        \vbox to \textheight{%
        \raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
        \vspace{2cm}\par
        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
        \hspace{1cm}\showjel\vss}}}
\def\blindtitle{{\sffamily
     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
        }}
\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}

\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
\let\maketitle\titlepage
\makeatother

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

%% KEYWORDS
\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}

%% ABSTRACT
\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
  {\end{minipage}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[showonlyrefs]{mathtools}
\usepackage[no-weekday]{eukdate}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

\wp{no/yr}
\jel{C10,C14,C22}

\author{Rob J~Hyndman, Yijun~Zeng, Han Lin~Shang}
\addresses{\textbf{Rob J Hyndman}\newline
Department of Econometrics \& Business Statistics\newline Monash
University, Clayton VIC 3800, Australia
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\newline Corresponding author\\[1cm]
\textbf{Yijun Zeng}\newline
Monash University
\\[1cm]
\textbf{Han Lin Shang}\newline
Australian National University
\\[1cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Hyndman, Zeng, Shang: \@date}
\makeatother

\usepackage{todonotes}

\begin{document}
\maketitle
\begin{abstract}
We forecast the old-age dependency ratio for Australia under various
pension age proposals, and estimate a pension age scheme that will
provide a stable old-age dependency ratio at a specified level. Our
approach involves a stochastic population forecasting method based on
coherent functional data models for mortality, fertility and net
migration, which we use to simulate future age-structures of the
population. We apply these ideas to Australian historical demographic
data and the results suggest that the pension age should be increased to
67 by 2021, 70 by 2032 and 75 by 2058 in order to maintain the old-age
dependency ratio at the current level, which is 21\%. This approach can
easily be extended to other target levels of the old-aged dependency
ratio and to other countries.
\end{abstract}
\begin{keywords}
functional time series model, coherent forecasts, demographic
components, pension and insurance
\end{keywords}

\section{Introduction}\label{introduction}

Australia's ageing population is a result of three factors: (1)
significant technological advances in medical care leading to longer
life spans; (2) high fertility rates after World War II resulting in the
``baby-boomer'' generation; and (3) a large decline in fertility rates
over the past twenty or so years \autocite{Fehr2008}. Consequently,
there are many older people leaving the workforce, and fewer workers
replacing them. Similar demographic challenges face many other
countries. One way of measuring this problem is via the old-age
dependency ratio (OADR): the number of people post-retirement to the
number of people of working age \autocite{Walker1982}. The OADR has been
increasing for many decades, but recently the rate of increase has
jumped markedly, and we show that it will continue to increase at a
rapid rate if there are no age-related policy changes.

This is consistent with the Australian Intergenerational Reports
\autocites{IGR02}{IGR07}{Swan2010} that explored the extent to which
existing age-related policies would affect the burden of future working
generations. To date, the Australian government has addressed the
problem through a variety of approaches including modifications to
compulsory superannuation policies, instigating work incentives for
older people, and increasing the pension age (defined as the age at
which eligible residents can access the age pension for the first time).
In 2009, the federal government announced an increase in pension age
from 65 to 67 by 2023 \autocite{Nielson2010}. In May 2014, the Federal
Treasurer proposed a further increase in the pension age to 70 by 2035.
This new proposal is in line with the recommendations from the
\textcite{PC13} and the \textcite{NCOA2014}. In this paper, we
investigate the effect of these changes in the pension age on the
forecast OADR.

@@

In this paper, we use the OADR as a measure of the financial burden on
the Australian workforce because a rise in the OADR will reduce the
ability of working age population to finance pensions and health costs
\autocites{ARW07}{IGR02}{IGR07}. The OADR is defined as the ratio of the
number of people aged over the pension age to the number of people of
working age \autocite{ARW07}. This can be expressed using the formula
below:

\begin{equation}
  \text{OADR} = \frac{\text{number of people aged over pension age}}{\text{number of people aged 15 to pension age}}*100\%
\end{equation}

We provide forecasts of future population age structures over the next
50 years. We then use those forecasts to compute the OADR associated
with pension age both before and after the adjustments. One contribution
of the current study is the combined use of the Hyndman-Ullah
forecasting method \autocite{HU07} and the product-ratio method
\autocite{HBY13} to forecast a population age structure. Forecasts of
population structure have been built on forecasts of age- and
gender-specific mortality rates, fertility rates and net migration using
this method. These forecasts presuppose that the historical trends of
past years will continue. Historical data of those rates have thus been
used as input to simulate future paths of age- and gender-specific
population structure.

Using these predicted population paths, we can calculate the average
OADR and predict the levels of old-age dependency that correspond to
different pension age schemes. By testing different pension age schemes,
we aim to find the pension age with the desired OADR outcome. Therefore,
other contributions of this research are (1) to examine the effect of
adjusting pension ages by comparing OADR before and after adjustment;
and (2) to suggest a suitable pension age scheme to ensure that the
economic burden will remain relatively constant at a desired level in
the future. It should be noted that the desired level has been set at
21\%, the estimated level of OADR in 2014. Whether this OADR level is
the most appropriate level for Australia remains to be examined and is
not the focus of this paper. This paper concentrates on providing a
framework for policy makers to find a target pension age for a given
desired level of OADR.

Some may argue that using OADR as a measure of financial burden for the
next generation could be biased, with a large number of retirees able to
rely on self-financed retirement income, such as superannuation.
However, 80\% of Australians over pension age rely fully or partly on
the age pension \autocite{Power14}. Therefore, the proportion of people
over working age is still a good representation of financial burden on
future working generations. Nevertheless, using OADR as a single
measurement of financial burden may overlook some important economic
factors, and \textcite{HY12} have suggested the use of an adjusted OADR
that takes account of some of these factors in analyzing pension
finance. We do not consider these here. Our approach is intended to
illustrate how population forecasting can be used to explore the pension
age issue in a simple context, and it could be extended to take into
account other variables.

The remainder of this paper proceeds as follows. Section
\ref{sec:litreview} reviews key literature on population forecasting and
the pension age in Australia. Section \ref{sec:data} describes the data
and modelling framework we use in this study. We show how our framework
can be used to estimate a pension age scheme that leads to a target OADR
in Section \ref{sec:target}. Finally, Section \ref{sec:conclusions}
gives concluding remarks as well as identifies the scope for future
research.

\section{Background and Literature Review}\label{sec:litreview}

\subsection{The age pension in
Australia}\label{the-age-pension-in-australia}

@@

Currently, the age pension is the primary source of income for 70\%
-80\% of Australian residents over 64 years \autocite{ABS06}. Even
though the government has promoted self-financed retirement by
introducing the Superannuation Guarantee in 1992, the Superannuation
Guarantee accumulations are still modest, with the current average less
than \$100,000 at retirement \autocite{ASFA07}. It has been estimated by
\textcite{PT08} that a retiree will need to have \$450,000 at 65 in
order to support a pension-equivalent income and this is more than eight
times of the average retirement accumulation of Australians. The
required wealth to comply with personal-funded retirement savings
requirements is too high for most Australians, suggesting that retirees
will continue to rely on the age pension in the coming decades.

The ageing trend in Australia, along with retirees' dependence on the
age pension, has dramatically increased the fiscal cost of the age
pension. For instance, the fiscal cost of the age pension is currently
around at 2.5\% of Gross Domestic Profit (GDP), but is expected to
increase to 4.4\% by the 2050s (Australian Government, 2007). The
\textcite{PC13} also indicates the government will need to raise taxes
by 21\% to pay for extra aged care and health costs, which is caused by
longer life expectancies in newer generations.

The issue is not only a problem for the government, but also for future
retirees. Given children born today are expected to live up to their 90s
on average, the \textcite{PC13} suggests that Australians won't be able
to afford to spend 35 years in retirement (i.e., retire around 65). They
have predicted that a third of baby boomers are expected to spend all
their money and assets before they
die\footnote{Baby Boomers are people born during the Post-World War II baby boom between the years 1946 and 1964}.
This problem might eventually deteriorate the sustainability of the age
pension, although the welfare system works well at the moment, because
demographic changes often have delayed effects \autocite{IGR02}.

\begin{figure}
\centering
\includegraphics{PensionAge_files/figure-latex/changespension-1.pdf}
\caption{\label{fig:changespension}Government's adjustments on pension age
scheme.}
\end{figure}

The Australian government is trying to take early intervention in
relation to the pension age to balance between providing social support
and fiscal cost. The pension age for males has been set at 65 for over a
century. While the pension age for females was 60 years until 1994, it
has steadily increased to 65 years, in line with that of males. The
previous Labor government announced in 2009 that the pension age will
increase six months every two years until 2017, capping at 67 in July
2023. This change is expected to have a significant impact on labor
force participation and to decrease the retirement duration of residents
because pension age is regarded as a normal age for retirement in
Australia. However, the \textcite{PC13} proposed further lifting of the
pension age to 70 on the grounds that increasing the pension age to 70
would save taxpayers \$150 billion in welfare costs and health spending,
and elderly people are capable of continuing work. The National
Commission of Audit (NCOA) (2014) holds the same opinion, and
recommended lifting the pension age to 70 from 2053. The current
Treasurer Joe Hockey announced that the government would increase the
pension age to 70 from 2035, which is proposed in the Federal Budget
Statement for 2014-2015 to be confirmed by Australian Federal
Government. Figure \ref{fig:changespension} illustrates the process of
the adjustments on pension age scheme. This announcement received
overwhelmingly negative responses in the media. \textcite{Guest14}
argues from an economic prospective that Australia currently has much
better circumstances than other OECD countries in terms of age pension
pressure, because spending on the age pension in Australia is less than
half of the OECD average. He also claims that Australia has a higher
consumption of all other goods and services, and so why are they not
allowed to have more leisure time. Further, \textcite{Power14} contends
that many Australians are not physically able to work full-time to 70,
so that recommendations made by the PC and NCOA are simplistic solutions
to a more a complex social problem.

In fact, all supportive arguments for increasing the pension age in
Australia have only justified the need to increase the pension age. No
justification on the pace or degree of the change has been provided.
Also, the government has not disclosed the method it used to determine
the new pension age. It is therefore not surprising that the adjustment
of the pension age has been questioned by the public. A consistent
method to determine the appropriate pension age would be more persuasive
in general. For example, it is preferred by OECD countries that pension
age should be linked to life expectancy \autocite{OECD12}. There is a
practical issue as to how to link life expectancy with pension ages
given that there is no information on future mortality rates
\autocite{PC13}. \textcite[p7]{CEDA07} proposed directly linking pension
age with life expectancy in the way that ``pension age should increase
by approximately 50\% of any increase in life expectancy''. However,
using a ``rule of thumb'' approach might not be ideal, as 50\% of the
change in life expectancy may not be a good adjustment to pension age in
the long term. Our study suggests an approach, linking the pension age
with the the aged dependency ratio, based on forecasts of future
mortality and fertility rates. Our approach has the advantage that it
can be used to justify the change in pension age and directly links the
pension age with our ultimate goal of a stable financial burden for the
public.

\subsection{Population forecasting}\label{population-forecasting}

In population forecasting, two very different approaches: a
deterministic approach (sometimes called a quasi-statistical approach)
and a stochastic approach (a purely statistical approach) are frequently
used. For instance, the Australian Bureau of Statistics (ABS) is using a
deterministic approach for its demographic projection, while some other
countries or institutes forecast demographic trends based on purely
statistical approaches such as Statistics New Zealand and the United
States Census Bureau. In the deterministic approach, experts project
future demographic changes using judgement-based scenarios. It ignores
the non-demographic factors such as major pandemics or wars, although
historically they are important factors that affect demographic
behaviour \autocite{PC13}. In stochastic approaches, the Lee-Carter (LC)
method \autocite{LC92} is the most prominent demographic forecasting
method, which now has several variants and extensions that relax some of
the assumptions of the basic Lee-Carter method \autocite[see,
e.g.,][]{SBH11}. Although statistical forecasting models are rigorous,
they still leave a basis for judgment-based adaptation, as they
sometimes produce unrealistic projections of demographic factors such as
long-run projections of mortality rates \autocite{PC13}.

Therefore, some official agencies decide to mix the two approaches in
their forecasts in order to provide a supplement to both approaches. For
instance, the most recent forecast of population in Australia by the
\textcite{PC13} was based on basic Lee-Carter methods plus some
judgement-based adaptations to address the characteristics of
demographic trends that the Lee-Carter method failed to capture. The
approach of forecasting used in this study is very similar to the
\textcite{PC13}. The Hyndman-Ullah method (2007) is used as a basic
forecasting approach, which is a comprehensive extension of Lee-Carter
method. In addition, an adapted product-ratio method is used to
eliminate the potential long-term divergence between female and male
mortality rates. A more detailed review of forecasting methods is given
below.

\subsubsection{Deterministic Methods Used by the
ABS}\label{deterministic-methods-used-by-the-abs}
